---
title: "Mistake Pitches"
author: "Justin Ford"
date: "2/8/2022"
output: 
  html_document:
    theme: "united"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, 
                      message = FALSE)
```

## What makes a mistake?

You know it when you see it: the batter's eyes light up, the pitcher's gaze drops to the ground just as the bat connects. A meatball. Even casual baseball fans probably have some familiarity with the concept of a so-called "mistake pitch." The breaking ball that doesn't *quite* break enough, the fastball right down Broadway. Ultimately, I think the simplest definition of a mistake pitch is a pitch that's not just likely to get hit, but likely to get *crushed.* Of course, not every pitch that gets crushed should be classified as a "mistake," nor is every inaccuracy punished by the hitter. Sometimes, a batter puts a good swing on a good pitch. Then how can we discern if a pitch is likely to get crushed or not? What makes a pitch a mistake pitch?

The first step is to find out what exactly "crushed" means. There are a few ways to do this: 

*   Set a threshold for run value, where all batted balls over the threshold are considered crushed. However, run values don't always correlate closely with how "well-hit" a batted ball is. Bloopers tend to have a high run value because they fall for hits often, but wouldn't be considered to be hit squarely or hard. 
*   Use Statcast's Barrel classification as a proxy for "crushed." This is the simplest and most direct option, because these are literally the batted balls that Statcast defines as "hit well." The one issue with using barrels is that batters will have widely differing abilities to hit barrels. Some hitters, like David Fletcher or Myles Straw, will rarely ever have a batted ball classified as a "Barrel," even if they hit it as well as they can. Does that mean it's impossible to throw a mistake pitch to David Fletcher? I don't think so. 
*   Use the subset of the "best" batted balls by each hitter. This solves the problem with using Statcast Barrels. However, I would need to devise a system to come up with evaluating with batted balls are the "best," or rather, "most crushed." The best way I could think to evaluate batted balls is by using run values or the built-in "estimated_woba_using_speedangle" column. Of course, this then presents us with the same problem we saw in option 1, where estimated_woba_using_speedangle isn't necessarily a measure of how well a ball is hit in the same way that Barrels are.

For simplicity's sake, I decided to just use Barrels as my stand-in for "crushed." It is the most direct measure of what I mean by "crushed" that I have readily available.

```{r libraries, include = FALSE}
source(file = "config.R")
library(tidyverse)
library(baseballr)
library(RMySQL)
library(gt)
library(xgboost)
library(Ckmeans.1d.dp)
library(MLmetrics)
library(mltools)
library(data.table)
library(SHAPforxgboost)

options(scipen = 999999)
```

So, looking at every pitch from the 2021 season (data retrieved using the excellent baseballr package for scraping from Baseball Savant), I constructed a model that would predict the probability that a given pitch would be barreled. 

```{r get_data, include = FALSE, cache = TRUE}
con <- dbConnect(MySQL(), dbname = dbname, 
                 user = user, 
                 password = password, 
                 host = host)


query <- "SELECT * FROM statcast_full WHERE game_year = 2021"

sc <- dbGetQuery(con, query)

dbDisconnect(con)
```

The features of my model included pitch velocity, location, movement, pitcher release point and the count. My model does not include any information regarding batter tendencies, or any contextual information about the pitcher's arsenal. It is simply looking on a pitch-by-pitch basis, regardless of batter, and trying to determine how often that pitch gets barreled on average.

```{r preprocessing, include = FALSE, cache = TRUE}

sc <- sc %>%
  select(-row_names, -spin_dir:-break_length_deprecated, -game_type, -hit_location, -game_year, -on_3b:-on_1b,
         -hc_x:-sv_id, -pitcher_1:-fielder_9)

# these are the "chances" (swings + called strikes)
# I want to train a model to predict the probability that a pitch will be swung at or
#   called a strike
chances <- c("called_strike", "foul", "foul_tip", "hit_into_play", 
             "swinging_strike", "swinging_strike_blocked")

# this step filters out all the bunt attempts that I can discern
# Since players who are bunting aren't attempting to barrel the ball, I throw these pitches out.
sc <- sc %>%
  filter(!(type == "X" & grepl(" bunt", des))) %>%
  filter(!(grepl("bunt", description))) %>%
  mutate(chance = ifelse(description %in% chances, 1, 0))

# this step takes features that measure in the x direction (horizontal movement/location)
# and flips their sign for lefties (i.e. negative values become positive.)
# This is done to make lefties pitches look like righties pitches to help the model train
# more accurately.
data <- sc %>%
  mutate(plate_x = ifelse(p_throws == "L", -1*plate_x, plate_x),
         pfx_x = ifelse(p_throws == "L", -1*pfx_x, pfx_x),
         release_pos_x = ifelse(p_throws == "L", -1*release_pos_x, release_pos_x)) %>%
  mutate_at(c("release_speed", "plate_x", "plate_z", "pfx_x", "pfx_z", "release_pos_x", "release_pos_z"),
            ~(scale(.) %>% as.vector))

# making some features
data <- data %>%
  mutate(barrel = ifelse(is.na(barrel), 0, barrel),
         same_hand = ifelse(p_throws == stand, 1, 0),
         stand = ifelse(stand == "R", 1, 0),
         balls = ifelse(balls == 4, 3, balls),
         count = factor(str_c(as.character(balls), as.character(strikes), sep = "-"))
         )

# one hot encoding categorical features that I want to include in the model
data <- one_hot(as.data.table(data), cols = "count")

```


```{r model, include = FALSE, cache = TRUE}
# Now that all of our data preprocessing is done, I have to split the train and test set.

train_size <- floor(0.75 * nrow(data))
set.seed(123)
train_ind <- sample(seq_len(nrow(data)), size = train_size)

train <- data[train_ind, ]
valid <- data[-train_ind, ]

# What we want to do is train two models:
#   - A model that predicts the probability of a pitch being barrelled.
#   - A model that predicts the probablitiy of a pitch being swung at or called for a strike.

xgb_data <- train %>%
  select(release_speed, same_hand, stand, plate_x, plate_z, pfx_x, pfx_z, release_pos_x, release_pos_z, 
         `count_0-0`, `count_1-0`, `count_2-0`, `count_3-0`, 
         `count_0-1`, `count_1-1`, `count_2-1`, `count_3-1`,
         `count_0-2`, `count_1-2`, `count_2-2`, `count_3-2`)

barrel_label <- train$barrel
chance_label <- train$chance

# run cv to find nrounds for models
barrel_cv <- xgb.cv(data = as.matrix(xgb_data), label = barrel_label, nrounds = 150, nfold = 5,
                    early_stopping_rounds = 50, metrics = "logloss", eta = 0.1, objective = "binary:logistic")

chance_cv <- xgb.cv(data = as.matrix(xgb_data), label = chance_label, nrounds = 250, nfold = 5,
                    metrics = "logloss", eta = 0.1, objective = "binary:logistic")

# getting the proper number of model iterations to prevent under/overfitting
eval_log <- as.data.frame(barrel_cv$evaluation_log)
nrounds_barrel <- which.min(eval_log$test_logloss_mean)

eval_log <- as.data.frame(chance_cv$evaluation_log)
nrounds_chance <- which.min(eval_log$test_logloss_mean)

# train models
barrel_mod <- xgboost(data = as.matrix(xgb_data), label = barrel_label, 
                   eta = 0.1, nrounds = nrounds_barrel, objective = "binary:logistic")

chance_mod <- xgboost(data = as.matrix(xgb_data), label = chance_label,
                      eta = 0.1, nrounds = nrounds_chance, objective = "binary:logistic")

xgb_valid <- valid %>%
  select(release_speed, same_hand, stand, plate_x, plate_z, pfx_x, pfx_z, release_pos_x, release_pos_z,
         `count_0-0`, `count_1-0`, `count_2-0`, `count_3-0`, 
         `count_0-1`, `count_1-1`, `count_2-1`, `count_3-1`,
         `count_0-2`, `count_1-2`, `count_2-2`, `count_3-2`)

# make preds on validation set
barrel_preds <- predict(barrel_mod, as.matrix(xgb_valid))
chance_preds <- predict(chance_mod, as.matrix(xgb_valid))

barrel_loss <- LogLoss(barrel_preds, valid$barrel)
chance_loss <- LogLoss(chance_preds, valid$chance)


# Compute feature importance matrix
barrel_importance <- xgb.importance(colnames(xgb_data), model = barrel_mod)
chance_importance <- xgb.importance(colnames(xgb_data), model = chance_mod)

```

After training the model, we can look at what features of the data contribute the most to correctly predicting whether the pitch was barreled or not.

```{r importance, cache=TRUE}

# Nice graph
xgb.ggplot.importance(barrel_importance[1:10,]) +
  theme_bw() +
  theme(
    legend.position = "none"
  ) +
  labs(
    title = "Feature Importance for XGBoost Model Predicting Barrels"
  )

```

A more complete description of what some of these attributes are measuring can be found [here](https://baseballsavant.mlb.com/csv-docs), but essentially plate_z and plate_x refer to vertical and horizontal location respectively. These are by far the most important features for them model. This is probably as expected, though I would've expected movement (pfx_x and pfx_z are the measurements for movement) to be a bigger contributor.

```{r predict, include = FALSE, cache=TRUE}

xgb.plot.importance(chance_importance[1:10,])

data <- data %>%
  select(release_speed, same_hand, stand, plate_x, plate_z, pfx_x, pfx_z, release_pos_x, release_pos_z,
         `count_0-0`, `count_1-0`, `count_2-0`, `count_3-0`, 
         `count_0-1`, `count_1-1`, `count_2-1`, `count_3-1`,
         `count_0-2`, `count_1-2`, `count_2-2`, `count_3-2`)

full_barrel_preds <- predict(barrel_mod, as.matrix(data))
full_chance_preds <- predict(chance_mod, as.matrix(data))

sc$xBarrel <- full_barrel_preds
sc$xChance <- full_chance_preds

```

We can bin pitches by their location and look at the average barrel probability for pitches in each bin.

```{r xbarrel_plot, cache=TRUE}

sc %>%
  filter(abs(plate_x) < 5) %>%
  ggplot(aes(plate_x, plate_z, z = xBarrel)) + 
  geom_tile(binwidth = .25, stat = "summary_2d", fun = mean, 
              na.rm = TRUE) +
  geom_rect(aes(xmin = -5/6, xmax = 5/6, ymin = 1.52, ymax = 3.67), color = "white", alpha = 0) +
  theme_bw() +
  coord_fixed() +
  labs(
    x = "Horizontal Pitch Location",
    y = "Vertical Pitch Location",
    title = "Barrel Probability by Pitch Location",
    subtitle = "Strikezone From Catcher's Perspective",
    fill = "Barrel Probability"
  ) +
  xlim(-10/6,10/6) +
  ylim(0, 5)



```
\
As we would expect, pitches nearest the middle of the zone have the highest probability of getting barreled according to the model.
\

## Which Pitchers Avoid Mistakes?

Now that we have a way to quantify how much of a mistake a pitch is (barrel probability, which I will also call xBarrels), we can look at the pitchers who are the best and worst at avoiding these pitches. However, I don't just want to look at who throws the most mistake pitches, I want to see which pitchers are able to prevent hard hits while also throwing pitches that get lots of swings and called strikes. A pitcher who throws 55 foot curveballs every pitch technically isn't throwing any mistake pitches by our definition, because those pitchers are essentially impossible to barrel. Those pitches, though, have little chance of accomplishing anything positive for the pitcher's team. So, I trained another model which attempted to predict the likelihood that a pitch would either be: a) swung at, or b) called for a strike. I call this likelihood "xChance" (i.e. the "chance" that the pitch results in a strike or batted ball). This way, I'm rewarding pitchers that are able to throw strikes *and* avoid mistakes. 

So we have a measure for the number of "mistakes" a pitcher made (xBarrels), and a measure for the number of "chances" that a pitcher allows for batters to barrel a pitch. We can simply take the proportion of xChances that end up as xBarrels for each pitcher, and we will have a good measure for determining pitchers that throw highly barrelable pitches:

```{r mistake_leaderboard}

chadwick_player_lu_table <- get_chadwick_lu()

chadwick_player_lu_table <- chadwick_player_lu_table %>%
  select(name_first, name_last, key_mlbam)

p_barrels <- sc %>%
  group_by(pitcher) %>%
  summarize(pitches = n(),
            chances = sum(chance),
            barrels = sum(barrel, na.rm = TRUE),
            xbarrels = sum(xBarrel),
            xchances = sum(xChance),
            barrel_rate = barrels/pitches,
            xbarrel_rate = xbarrels / pitches,
            xbpc = xbarrels/xchances) %>%
  left_join(chadwick_player_lu_table, by = c("pitcher" = "key_mlbam")) %>%
  mutate(name = str_c(name_first, name_last, sep = " ")) %>%
  select(-pitcher, -name_first, -name_last)

p_barrels %>%
  filter(pitches > 1500) %>%
  arrange(desc(xbpc)) %>%
  select(-barrel_rate, -xbarrel_rate) %>%
  select(name, everything()) %>%
  head(10) %>%
  gt() %>%
  tab_header(
    title = "Top 10 Barrel Prone Pitchers, 2021 MLB Season",
    subtitle = "1500 pitch min."
  ) %>%
  tab_options(
    heading.title.font.weight = "bold",
    table.width = px(700)
  ) %>%
  opt_row_striping() %>%
  cols_label(
    name = "Pitcher",
    pitches = "Pitches",
    chances = "Chances",
    barrels = "Barrels",
    xbarrels = "xBarrels (xB)",
    xchances = "xChances (xC)",
    xbpc = "xB/xC"
  ) %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(columns = xbpc, decimals = 4) %>%
  fmt_number(columns = c(xbarrels, xchances), decimals = 0)

```

\
And the pitchers who are the best at throwing hard-to-barrel pitches:

\
```{r mistake_leaderboard2}

p_barrels %>%
  filter(pitches > 1500) %>%
  arrange((xbpc)) %>%
  select(-barrel_rate, -xbarrel_rate) %>%
  select(name, everything()) %>%
  head(10) %>%
  gt() %>%
  tab_header(
    title = md("Top 10 *Least* Barrel Prone Pitchers, 2021 MLB Season"),
    subtitle = "1500 pitch min."
  ) %>%
  tab_options(
    heading.title.font.weight = "bold",
    table.width = px(700)
  ) %>%
  opt_row_striping() %>%
  cols_label(
    name = "Pitcher",
    pitches = "Pitches",
    chances = "Chances",
    barrels = "Barrels",
    xbarrels = "xBarrels (xB)",
    xchances = "xChances (xC)",
    xbpc = "xB/xC"
  ) %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(columns = xbpc, decimals = 4) %>%
  fmt_number(columns = xbarrels, decimals = 2)

```
\

## Pitcher Mistakes

Since we have the probability that every pitch will be barreled, we can plot the pitches with the highest barrel probability (these are our "mistakes"). I'm taking the top 1.5% of pitches to be our "mistakes," since this is roughly the percentage of pitches that are barreled league-wide.


```{r mistake_plot, fig.asp=1, fig_width = 15, echo=FALSE}

# "mistakes" plot
sc %>%
  arrange(desc(xBarrel)) %>%
  mutate(pitch_name = ifelse(pitch_name == "Fastball", "4-Seam Fastball", pitch_name)) %>%
  filter(row_number() < (nrow(sc) * 0.015)) %>%
  ggplot(aes(plate_x, plate_z)) +
  geom_point(size = 4, alpha = 0.3, aes(color = stand)) +
  geom_segment(x = -5/6, y = 3.67, xend = -5/6 , yend = 1.52) + # draw strikezone
  geom_segment(x = 5/6, y = 3.67, xend = 5/6 , yend = 1.52) +
  geom_segment(x = -5/6, y = 3.67, xend = 5/6 , yend = 3.67) +
  geom_segment(x = -5/6, y = 1.52, xend = 5/6 , yend = 1.52) +
  coord_fixed() +
  theme_bw() +
  xlim(-1,1) +
  ylim(1.5,3.75) +
  facet_wrap(~pitch_name) +
  labs(
    title = "Mistake Pitches By Pitch Type, 2021 MLB Season",
    subtitle = "Catcher's Perspective",
    x = "Horizontal Plate Location",
    y = "Vertical Plate Location"
  ) +
  scale_color_discrete(name = "Batter Side") +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())


```
\
Once again, we see how many mistakes are pitches that are left over the middle of the plate, for all pitch types. We can also see how more mistakes tend to be made on the inside part of the plate. Also, breaking balls and offspeed pitches up in the zone don't really show up here as "mistakes," as maybe conventional wisdom would suggest. 

